TTS2016R: A data set to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Golden Horseshoe Area, Ontario, Canada

### 2022-10-24 Manuscript-Data-Package/Manuscript-Data-Package.tex
submitted revision to journal. submission package in 'editorial-correspondance'.

\author{Anastasia Soukhov\affilnum{}, Antonio Páez\affilnum{}}

@@

\begin{abstract}
This paper describes and visualises the data contained within the
\{TTS2016R\} data package created in \texttt{R}, the statistical
computing and graphics language. In addition to a synthetic example,
\{TTS2016R\} contains home-to-work commute information for the Greater
Golden Horseshoe (GGH) area in Canada retrieved from the 2016
Transportation Tomorrow Survey (TTS). Included are all Traffic Analysis
Zones (TAZ), the number of people who are employed full-time per TAZ,
the number of jobs per TAZ, the count of origin destination (OD) pairs
and trips by mode per origin TAZ, calculated car travel time from TAZ OD
centroid pairs, and associated spatial boundaries to link TAZ to the
Canadian Census. To illustrate how this information can be analysed to
understand patterns in commuting, we estimate a distance-decay curve
(i.e., impedance function) for the region. The value in \{TTS2016R\} is
that it is a growing open data product built on \texttt{R}
infrastructure that allows for the immediate access of home-to-work
commuting data alongside complimentary objects from different sources.
The package will continue expanding with additions by the authors and
the community at-large by requests in the future. \{TTS2016R\} can be
freely explored and downloaded in the associated
\href{https://github.com/soukhova/TTS2016R}{Github repository} where the
documentation and code involved in data creation (including this
manuscript), manipulation, and all open data products are detailed.
\end{abstract}

\keywords{Jobs; population; work; commute; travel time; impedance;
Greater Toronto and Hamilton Area; Ontario, Canada; R}

@@

In the GGH, there are 3,446,957 workers, 3,081,900 jobs, and 3,282,611
work-related trips (for the 2016 TTS survey day). The values are
organized within the origin destination (OD) table in the \{TTS2016R\}
package and are derived from the cross-tabulation by person and by trip
for the full-time employed population and associated places of
employment. The TTS is a proportionally representative survey, hence the
values included in \{TTS2016R\} are adjusted to reflect the GGH
population.

It is important to note that the total number of full-time workers and
jobs in the TTS 2016 region are not equal. Since the outer boundaries of
the TTS are permeable, workers who reside within the boundaries but have
workplaces that are outside of the boundaries are counted as workers
within an origin TAZ, while jobs in TAZ that are filled by workers who
reside outside the GGH boundaries are \emph{unknown} since they were not
surveyed. This mismatch results in the total number of workers being
1.12 times larger than the number of jobs (i.e., 3,446,957 workers to
3,081,900 jobs). As such, the OD table contained in \{TTS2016R\} offers
a perspective on all workers in the GGH and their home-based trips to
places of GGH employment.

@@

Figure \ref{fig:tts-workers-jobs-plot} presents the number of employed
people and associated jobs per TAZ. It can be observed that the spatial
distribution of jobs and workers is unequal, which is indicative of a
jobs -housing imbalance that can impact accessibility in a region
\citep{Levine1998rethinking}. It can also be seen that there is a higher
number of TAZ with no workers than zones with no jobs (i.e., 791 TAZ
with no workers : 396 TAZ with no jobs) and the mean of workers per TAZ
is higher than the mean of jobs. The number TAZ with an extreme number
of jobs at the highest and lowest percentiles is significantly higher
than the number of workers.

@@

Also included in \{TTS2016R\} are the estimated travel times between OD
as summarized in descriptive statistics table in Figure
\ref{fig:plot-tt-ttpertrip}; travel times are calculated using the
package \{r5r\}. \{r5r\} interfaces with the java-based R5 routing
engine developed separately by Conveyal
\citep{conveyalConveyalR5Routing2022}. The inputs to \{r5r\} for this
data package were the desired mode, a maximum travel time threshold of
180 minutes, the geo-coded origin destination pairs based on the
centroids of the TAZ, and the static Open Street Map road network of
Ontario (retrieved using Geofabrick
\citep{geofabrikOntarioOpenStreetMapGeofabrik2022}). A travel time
threshold of 180 minutes was selected since it captures almost all
potential OD interactions.

Additionally, the car mode was included since it is a critically
important commute mode in the GGH. 2,598,379 of the trips are made using
a car mode out of the total 3,282,611 work-related trips according to
the TTS 2016 data (i.e., 79\% of trips are taken by car).

These travel times are useful addition to \{TTS2016R\} since they are
not included in the TTS Data Retrieval System but they are vitally
important to estimate the cost of travel and associated impedance
functions, among other possible applications. If the readership is
interested in additional information regarding the travel time
computation, please see the calculation notebook in the documentation of
\{TTS2016R\} and details about \{r5r\} at the
\href{https://ipeagit.github.io/r5r/index.html}{r5r package website}.

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS. Planning boundaries of Niagara and Waterloo (Data Management Group, 2018a), and the Toronto census metropolitan area (Statistics Canada, 2017) are drawn with purple, brown and blue borders, respectively. }\label{fig:plot-tt-ttpertrip}
\end{figure}
\newpage

As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total
travel time resembles the spatial trend distribution in the number of
employed people in the previous plot (Figure
\ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the
average travel time is distinct from other plots presented so far. For
instance, we can see that in areas around the south-eastern border such
as Niagara and Waterloo (purple and brown borders), the average travel
times are moderately low. Additionally, travel times (by car) within the
core of the Toronto census metropolitan area (CMA) (blue) is also
moderately since traffic congestion is not reflected in the travel time
estimations. Further from these areas, travel times are higher.

@@

Impedance functions are useful to understand mobility behaviour and are
used to estimate gravity models of spatial interaction
\citep{wilson1971, haynes_gravity_1985} and applied in accessibility
analysis
\citep{hansen_how_1959, talen_assessing_1998, paez_jobs_2013, barboza_balancing_2021}.
An impedance function \(f(\cdot)\) depends on the cost of travel
\(c_{ij}\) between locations \(i\) and \(j\) (all which is supplied in
the travel time and origin-destination table within \{TTS2016R\}).

@@

\begin{equation}
\label{gamma-dist}
\begin{array}{l}\ 
f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty\\
\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
\end{array}
\end{equation}\textbackslash end\{equation\}

\newpage

@@

\{TTS2016R\}, the open data package introduced in this paper fuses
multiple sources of data. It includes an OD cross-tabulation by person
and by trip mode table for home-to-work commute data from the 2016 TTS
alongside complimentary boundaries and estimated travel times. The value
of this data package is in its transparency, easy of access, and its
open infrastructure for the addition of complimentary data sets in the
future. Using \texttt{R} users can immediately and easily explore GGH
commute flow trends as well as suggestion further amendments to the
package by request. One possible use of this data, as showcased in this
paper, is the calibration of impedance functions which in turn can be
used for accessibility analysis.

@@

\bibliographystyle{sageh}
\bibliography{bibfile}

>>>>

\author{Anon1\affilnum{}, Anon2\affilnum{}}

@@

\begin{abstract}
This paper describes and visualises the data contained within the
\{TTS2016R\} data package created in \texttt{R}, the statistical
computing and graphics language. In addition to a synthetic example,
\{TTS2016R\} contains home-to-work commute information for the Greater
Golden Horseshoe (GGH) area in Canada retrieved from the 2016
Transportation Tomorrow Survey (TTS). Included are all Traffic Analysis
Zones (TAZ), the number of people who are employed full-time per TAZ,
the number of jobs per TAZ, the count of origin destination (OD) pairs
and trips by mode per origin TAZ, calculated car travel time from TAZ OD
centroid pairs, and associated spatial boundaries to link TAZ to the
Canadian Census. To illustrate how this information can be analysed to
understand patterns in commuting, we estimate a distance-decay curve
(i.e., impedance function) for the region. \{TTS2016R\} is a growing
open data product built on \texttt{R} infrastructure that allows for the
immediate access of home-to-work commuting data alongside complimentary
objects from different sources. The package will continue expanding with
additions by the authors and the community at-large by requests in the
future. \{TTS2016R\} can be freely explored and downloaded in the
associated \href{https://github.com/soukhova/TTS2016R}{Github
repository} where the documentation and code involved in data creation,
manipulation, and all open data products are detailed.
\end{abstract}

\keywords{Jobs; population; work; commute; travel time; impedance;
Greater Toronto and Hamilton Area; Greater Golden Horshoe Area, Ontario,
Canada; R}

@@

In the GGH, there are 3,446,957 workers, 3,081,900 jobs, and 3,282,611
work-related trips (for the 2016 TTS survey day). The values are
organized within the origin destination (OD) table in the \{TTS2016R\}
package and are derived from the cross-tabulation by person and by trip
for the full-time employed population and associated places of
employment.

Additionally, the total number of full-time workers and jobs in the TTS
2016 region are not equal. Since the outer boundaries of the TTS are
permeable, workers who reside within the boundaries but have workplaces
that are outside of the boundaries are counted as workers within an
origin TAZ, while jobs in TAZ that are filled by workers who reside
outside the GGH boundaries are \emph{unknown} since they were not
surveyed. This mismatch results in the total number of workers being
1.12 times larger than the number of jobs. The TTS is a proportionally
representative survey, hence the values included in \{TTS2016R\} are
adjusted to reflect the GGH working population and their home-based
trips to places of GGH employment.

@@

Figure \ref{fig:tts-workers-jobs-plot} presents the number of employed
people and associated jobs per TAZ. It can be observed that the spatial
distribution of jobs and workers is unequal, which is indicative of a
jobs -housing imbalance that can impact accessibility in a region
\citep{Levine1998rethinking}. Also, there is a higher number of TAZ with
no workers than zones with no jobs (i.e., 791 TAZ with no workers and
396 TAZ with no jobs) and the mean of workers per TAZ is higher than the
mean of jobs. The number TAZ with an extreme number of jobs at the
highest and lowest percentiles is significantly higher than the number
of workers.

@@

Also included in \{TTS2016R\} are the estimated travel times between OD
as summarized in the descriptive statistics table in Figure
\ref{fig:plot-tt-ttpertrip}; travel times are calculated using the
package \{r5r\}. \{r5r\} interfaces with the java-based R5 routing
engine developed separately by Conveyal
\citep{conveyalConveyalR5Routing2022}. The inputs to \{r5r\} for this
data package were: the desired mode, a maximum travel time threshold of
180 minutes, the geo-coded origin destination pairs based on the
centroids of the TAZ, and the static OpenStreetMap road network of
Ontario (retrieved using Geofabrick
\citep{geofabrikOntarioOpenStreetMapGeofabrik2022}). A travel time
threshold of 180 minutes was selected since it captures almost all
potential OD interactions.

Additionally, car travel is included in this data package because it is
a critically important commute mode in the GGH. 2,598,379 of the trips
are made using a car out of the total 3,282,611 work-related trips
according to the TTS 2016 data (i.e., 79\% of trips are taken by car).

These travel times are a useful addition to \{TTS2016R\} since they are
not included in the \href{https://dmg.utoronto.ca/idrs/index}{TTS Data
Retrieval System} but they are vitally important to estimate the cost of
travel and associated impedance functions, among other possible
applications. If the readership is interested in additional information
regarding the travel time computation, please see the calculation
notebook in the documentation of \{TTS2016R\} and details about \{r5r\}
at the \href{https://ipeagit.github.io/r5r/index.html}{r5r package
website}.

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time by car (left) and average worker travel time by car (right) for each TAZ in the 2016 TTS. Car travel times calculated using {r5r} (Pereira et al., 2021). Planning boundaries of Niagara and Waterloo (Data Management Group, 2018a), and the Toronto census metropolitan area (Statistics Canada, 2017) are drawn with purple, brown and blue borders, respectively. }\label{fig:plot-tt-ttpertrip}
\end{figure}
\newpage

As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total
travel time resembles the spatial trend distribution in the number of
employed people in the previous plot (Figure
\ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the
average travel time is distinct from other plots presented so far. For
instance, we can see that in areas around the south-eastern border such
as Niagara and Waterloo (purple and brown borders), the average travel
times are moderately low. Additionally, travel times (by car) within the
core of the Toronto census metropolitan area (CMA) (blue) is also
moderately low since traffic congestion is not reflected in the travel
time calculations. Further from these areas, travel times are higher.

@@

An application of the \{TTS2016R\} package is the calculation of an
impedance function. Impedance functions are useful to understand
mobility behaviour and are used to estimate gravity models of spatial
interaction \citep{wilson1971, haynes_gravity_1985} and applied in
accessibility analysis
\citep{hansen_how_1959, talen_assessing_1998, paez_jobs_2013, barboza_balancing_2021}.
An impedance function \(f(\cdot)\) depends on the cost of travel
\(c_{ij}\) between locations \(i\) and \(j\) (all which is supplied in
the travel time and origin-destination table within \{TTS2016R\}).

@@

\renewcommand{\[}{\begin{equation}}
\renewcommand{\]}{\end{equation}}

\[
\begin{aligned}
f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty \label{gamma-dist} \\
\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx
\end{aligned}
\]

@@

This paper introduces \{TTS2016R\}, an open data product in the form of
a \texttt{R} data package. This package is a fusion of data from
multiple sources and we demonstrate the spatial and numeric extent of
the data contained within. It includes an OD cross-tabulation by person
and by trip mode table for home-to-work commute data from the 2016 TTS
alongside complimentary boundaries and estimated travel times. The value
of this data package is in its transparency, easy of access, and its
open infrastructure for the addition of complimentary data sets in the
future. \texttt{R} users can immediately and easily explore GGH commute
flow trends as well as suggestion further amendments to the package by
request. One possible use of this data, as showcased in this paper, is
the calibration of impedance functions which in turn can be used for
accessibility analysis.

@@

\bibliographystyle{sageh}
\bibliography{bibfile.bib}
